

```{r}
library(caret)
library(glmnet)
library(dplyr)
library(MASS)
library(ggplot2)
library(tidyverse)
library(reshape2)
library(pROC)

# Import the Dataset 
data <- read.csv("C:/SCHOOL Spring '25/STA4241/Project/Data/alzheimers_prediction_dataset.csv", sep = ',')

data
```

### Data Cleaning
```{r}
levels <- c("Low", "Medium", "High")

l_m_h_factors <- c("Physical.Activity.Level", "Depression.Level", "Air.Pollution.Exposure", "Income.Level", "Social.Engagement.Level", "Stress.Levels")
for (col in l_m_h_factors) {
  data[[col]] <- factor(data[[col]], levels = levels)
}

data$Alcohol.Consumption <- factor(data$Alcohol.Consumption, levels = c("Never", "Occasionally", "Regularly"))
data$Smoking.Status <- factor(data$Smoking.Status, levels = c("Never", "Former", "Current"))
data$Dietary.Habits <- factor(data$Dietary.Habits, levels = c("Unhealthy", "Average", "Healthy"))


for (col in names(data)) {
  if (is.character(data[[col]])) {
    data[[col]] <- factor(data[[col]])
  }
}

#Create a new data set with all numeric values
new_data <- data.frame()

factor_list <- c("Physical.Activity.Level", "Depression.Level", "Air.Pollution.Exposure", "Income.Level", "Social.Engagement.Level", "Stress.Levels",
                 "Alcohol.Consumption", "Smoking.Status", "Dietary.Habits", "Country", "Gender", "Diabetes", "Hypertension", "Cholesterol.Level",
                 "Family.History.of.Alzheimer.s", "Sleep.Quality", "Employment.Status", "Marital.Status", "Genetic.Risk.Factor..APOE.ε4.allele.", 
                 "Urban.vs.Rural.Living", "Alzheimer.s.Diagnosis")

# Convert the specified columns to numeric and combine them into a data frame
new_data <- as.data.frame(lapply(factor_list, function(col) as.numeric(data[[col]])))

# Optionally, set the column names correctly
names(new_data) <- factor_list

head(data)


```


### 2 Exploratory Data Analysis (EDA)
The description of all the involved variables are given below.

```{r}
# Extract numeric columns from 'data'
numeric_data <- data[, sapply(data, is.numeric)]

# Assuming new_data is already defined (even as an empty data frame with the same number of rows)
new_data <- cbind(new_data, numeric_data)

```
 Data Preparation/Cleaning

This exercise is to perform data 
processing/cleaning via the following steps. 

\begin{itemize}
\item 
Removing questionable observations and variables 
(i.e., any possibility of target leakage) 
that are not suitable as predictors for \verb|clm|. 
Explain your decisions.

\item 
Changing \verb|veh_age| and \verb|ageeat| to factor variables.
Discuss the advantage(s) and disadvantage(s) of the factor 
conversions in the context of GLMs. 
Regardless, convert the two variables to factors.
\end{itemize}

```{r}
summary(data)
dim(data)
```
## EDA

First, let's analyze the data and remove any variables that are not suitable as predictors for Alzheimer.s.Diagnosis due to target leakage. 

```{r}
head(data)
summary(data$Alzheimer.s.Diagnosis)
sum(is.na(data))
```
There doesn't seem to be any variables that directly leak the Alzheimer Diagnosis. However, there are some variables that will potentially cause high collinearity, like Dietary Habits and Diabetes. Let's do some more exploratory data analysis to see if these are correlated.

After viewing the summary of the data, it seems that all of the variables are normal and there are no missing values. It seems as though this dataset was cleaned prior to being uploaded to Kaggle, which makes our job for Data Preparation easy! However, we will still do our due dilligence in finding any potential interaction and/or collinearity.

Before looking at correlation/interaction, though, let's do some univariate data exploration to see whether any of our non-factor variables need to be transformed, and if so, how?

```{r}
ggplot(data, aes(x = data$Age)) + geom_histogram()
ggplot(data, aes(x = data$BMI)) + geom_histogram()
ggplot(data, aes(x = data$Education.Level)) + geom_histogram()
ggplot(data, aes(x = data$Cognitive.Test.Score)) + geom_histogram()


ggplot(data, aes(x = log(data$Age))) + geom_histogram(color="blue")
ggplot(data, aes(x = log(data$BMI))) + geom_histogram(color="blue")
ggplot(data, aes(x = log(data$Education.Level))) + geom_histogram(color="blue")
ggplot(data, aes(x = log(data$Cognitive.Test.Score))) + geom_histogram(color="blue")


ggplot(data, aes(x = sqrt(data$Age))) + geom_histogram(color="red")
ggplot(data, aes(x = sqrt(data$BMI))) + geom_histogram(color="red")
ggplot(data, aes(x = sqrt(data$Cognitive.Test.Score))) + geom_histogram(color="red")


ggplot(data, aes(x = 1/(data$Age))) + geom_histogram(color="yellow")
ggplot(data, aes(x = 1/(data$BMI))) + geom_histogram(color="yellow")
ggplot(data, aes(x = 1/(data$Cognitive.Test.Score))) + geom_histogram(color="yellow")

ggplot(data, aes(x = sqrt(1/(data$Age)))) + geom_histogram(color="orange")
ggplot(data, aes(x = sqrt(1/(data$BMI)))) + geom_histogram(color="orange")
ggplot(data, aes(x = sqrt(1/(data$Cognitive.Test.Score)))) + geom_histogram(color="orange")


ggplot(data, aes(x = log(1/(data$Age)))) + geom_histogram(color="green")
ggplot(data, aes(x = log(1/(data$BMI)))) + geom_histogram(color="green")
ggplot(data, aes(x = log(1/(data$Cognitive.Test.Score)))) + geom_histogram(color="green")

```

After looking at many possible transformations, it seems that perhaps it may be worth tranforming the Cognitive.Test.Score to log(1/(data$Cognitive.Test.Score)). Otherwise, it is probably best to leave the others as is since they are not becoming more normal.

Next, let's perform some Bivariate Data Exploration. Let's look at some boxplots comparing our numeric variables to our target variable.

```{r}
ggplot(data, aes(x=Alzheimer.s.Diagnosis, y=Age)) + geom_boxplot()
ggplot(data, aes(x=Alzheimer.s.Diagnosis, y=BMI)) + geom_boxplot()
ggplot(data, aes(x=Alzheimer.s.Diagnosis, y=Education.Level)) + geom_boxplot()
ggplot(data, aes(x=Alzheimer.s.Diagnosis, y=Cognitive.Test.Score)) + geom_boxplot()

```
Wow! It looks like Age plays a big factor in Alzheimer's Diagnoses (which we expect).

Looking at the means and medians of each of these:

```{r}
data %>%
  group_by_("Alzheimer.s.Diagnosis") %>%
  summarise(
    mean = mean(Age),
    median = median(Age),
    n = n()
  )
data %>%
  group_by_("Alzheimer.s.Diagnosis") %>%
  summarise(
    mean = mean(BMI),
    median = median(BMI),
    n = n()
  )
data %>%
  group_by_("Alzheimer.s.Diagnosis") %>%
  summarise(
    mean = mean(Education.Level),
    median = median(Education.Level),
    n = n()
  )
data %>%
  group_by_("Alzheimer.s.Diagnosis") %>%
  summarise(
    mean = mean(Cognitive.Test.Score),
    median = median(Cognitive.Test.Score),
    n = n()
  )

```
As expected, the means for each are very similar except for Age, which is significantly different.


To perform data exploration on the factor variables, let's create bar charts. We can color code them by Alzheimer's Diagnoses.
```{r}

# Select only categorical variables
cat.data <- data[, sapply(data, is.factor)]

# Loop through categorical columns
for (i in colnames(cat.data)) {
  plot <- ggplot(data, aes(x = .data[[i]], fill = factor(Alzheimer.s.Diagnosis))) + # Select Age as what to compare each to
    geom_bar(position = "fill") +
    labs(x = i, y = "percent") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
  print(plot)  # Print each plot
}
```
As we can see, the Family History of Alzheimer's and Genetic Risk Factor (APOE 4 allele) are the variables with the strongest correlation to the diagnoses since these are the plots who's proportions of negative diagnosis (in red) to positive (in blue) vary most significantly across the their levels.

Let's look at the mean of Alzheimer's Diagnosis by different levels of predictors to see more. (come back to this)

```{r}

test1 <- data[, sapply(data, is.factor)]

# Convert Alzheimer.s.Diagnosis to numeric
test1$Alzheimer.s.Diagnosis <- as.numeric(test1$Alzheimer.s.Diagnosis) - 1

# Perform the group_by and summarise operations
x <- test1 %>%
  group_by(Country) %>%
  summarise(mean = mean(Alzheimer.s.Diagnosis, na.rm = TRUE), n = n())

print(x)

```

Country has many different levels. Let's consider what would happen if we combined multiple countries into continents. Or what would happen if we grouped all countries with a mean of >0.48 together.

```{r}
levels(data$Country)

table(data$Country)

test2 <- data[, sapply(data, is.factor)]

levels(test2$Country)[levels(test2$Country) == "Brazil"] <- "South America"
levels(test2$Country)[levels(test2$Country) == "Argentina"] <- "South America"

levels(test2$Country)[levels(test2$Country) == "Canada"] <- "North America"
levels(test2$Country)[levels(test2$Country) == "USA"] <- "North America"
levels(test2$Country)[levels(test2$Country) == "Mexico"] <- "North America"

levels(test2$Country)[levels(test2$Country) == "China"] <- "Asia"
levels(test2$Country)[levels(test2$Country) == "India"] <- "Asia"
levels(test2$Country)[levels(test2$Country) == "Japan"] <- "Asia"
levels(test2$Country)[levels(test2$Country) == "Saudi Arabia"] <- "Asia"
levels(test2$Country)[levels(test2$Country) == "South Korea"] <- "Asia"

levels(test2$Country)[levels(test2$Country) == "France"] <- "Europe"
levels(test2$Country)[levels(test2$Country) == "Germay"] <- "Europe"
levels(test2$Country)[levels(test2$Country) == "Italy"] <- "Europe"
levels(test2$Country)[levels(test2$Country) == "Norway"] <- "Europe"
levels(test2$Country)[levels(test2$Country) == "Sweden"] <- "Europe"
levels(test2$Country)[levels(test2$Country) == "Russia"] <- "Europe"
levels(test2$Country)[levels(test2$Country) == "Spain"] <- "Europe"
levels(test2$Country)[levels(test2$Country) == "UK"] <- "Europe"


# Convert Alzheimer.s.Diagnosis to numeric
test2$Alzheimer.s.Diagnosis <- as.numeric(test1$Alzheimer.s.Diagnosis)

# Perform the group_by and summarise operations
x <- test2 %>%
  group_by(Country) %>%
  summarise(mean = mean(Alzheimer.s.Diagnosis, na.rm = TRUE), n = n())

print(x)


test3 <- test1

levels(test3$Country)[levels(test3$Country) == "Russia"] <- "Russia-India-Brazil-Mexico-SouthAfrica"
levels(test3$Country)[levels(test3$Country) == "India"] <- "Russia-India-Brazil-Mexico-SouthAfrica"
levels(test3$Country)[levels(test3$Country) == "Brazil"] <- "Russia-India-Brazil-Mexico-SouthAfrica"
levels(test3$Country)[levels(test3$Country) == "Mexico"] <- "Russia-India-Brazil-Mexico-SouthAfrica"
levels(test3$Country)[levels(test3$Country) == "South Africa"] <- "Russia-India-Brazil-Mexico-SouthAfrica"

levels(test3$Country)[!levels(test3$Country) == "Russia-India-Brazil-Mexico-SouthAfrica"] <- "Other"

# Perform the group_by and summarise operations
x <- test3 %>%
  group_by(Country) %>%
  summarise(mean = mean(Alzheimer.s.Diagnosis, na.rm = TRUE), n = n())

print(x)

```

Let's try and see if there is any multicollinearity that exists in the data.
First, let's try to look at the correlation matrix for only the numerical data.

```{r}
cor.matrix.numerical <- cor(data[, c(2, 4, 5, 13)])

round(cor.matrix.numerical, digits = 4)

```
There doesn't seem to be any meaningful correlation between these variables.


Now, let's look at the correlation between all of the data.

```{r}
cor.matrix.all <- cor(new_data)
round(cor.matrix.all, digits = 2)
melted.cor.matrix <- melt(round(cor.matrix.all, digits = 4))
ggplot(data = melted.cor.matrix, aes(x=Var1, y=Var2, fill=value)) + geom_raster()

```
There doesn't seem to be much correlation, which is good! 

Notice that Alzheimer.s.Diagnosis is correlated with some variables (which is good, since this will be our target variable). All other correlation is negligable.


```{r}
test4 <- data

levels(test4$Country)[levels(test4$Country) == "Russia"] <- "Russia-India-Brazil-Mexico-SouthAfrica"
levels(test4$Country)[levels(test4$Country) == "India"] <- "Russia-India-Brazil-Mexico-SouthAfrica"
levels(test4$Country)[levels(test4$Country) == "Brazil"] <- "Russia-India-Brazil-Mexico-SouthAfrica"
levels(test4$Country)[levels(test4$Country) == "Mexico"] <- "Russia-India-Brazil-Mexico-SouthAfrica"
levels(test4$Country)[levels(test4$Country) == "South Africa"] <- "Russia-India-Brazil-Mexico-SouthAfrica"

levels(test4$Country)[!levels(test4$Country) == "Russia-India-Brazil-Mexico-SouthAfrica"] <- "Other"

# Loop through categorical columns
for (i in colnames(test4)) {
  plot <- ggplot(test4, aes(x = .data[[i]], fill = factor(Age))) + # Select Age as what to compare each to
    geom_bar(position = "fill") +
    labs(x = i, y = "percent") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
  print(plot)  # Print each plot
  
  plot <- ggplot(data, aes(x = .data[["Country"]], fill = factor(Genetic.Risk.Factor..APOE.ε4.allele.))) +
  geom_bar(position = "fill") +
  labs(x = "Country", y = "percent") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
  print(plot) 
  
  plot <- ggplot(data, aes(x = .data[["Country"]], fill = factor(Age))) + # Select Age as what to compare each to
  geom_bar(position = "fill") +
  labs(x = "Country", y = "percent") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
  print(plot)
}


# Loop through categorical columns
for (i in colnames(test4)) {
  plot <- ggplot(test4, aes(x = .data[[i]], fill = factor(Genetic.Risk.Factor..APOE.ε4.allele.))) +
    geom_bar(position = "fill") +
    labs(x = i, y = "percent") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
  
  print(plot)  # Print each plot
}



data <- test4
```
By looking at the plots, it seems as though country may be potentially correlated with a genetic risk factor and/or Age (prior to combining). Therefore, 

Let's now revisit the question of whether or not Diabetes and Diet interact. Let's generate the bar chart for Diabetes filled by the target, Alzheimer's Diagnosis, and faceted by Diet.
```{r}
# Ensure 'Alzheimer.s.Diagnosis' and 'Diabetes' are factors
data$Alzheimer.s.Diagnosis <- as.factor(data$Alzheimer.s.Diagnosis)
data$Diabetes <- as.factor(data$Diabetes)

# Create the plot
ggplot(data, aes(x = Diabetes, fill = Alzheimer.s.Diagnosis)) +
  geom_bar(position = "fill") + 
  facet_wrap(~ Dietary.Habits) +
  labs(y = "Proportion", x = "Diabetes") +
  theme_minimal()
```
It seems that the charts are generally even, so these do not interact. How about BMI and Diet? BMI and Physical Activity level?

```{r}
ggplot(data, aes( x = Alzheimer.s.Diagnosis, y = BMI)) + geom_boxplot() + facet_wrap(~ Dietary.Habits)

ggplot(data, aes( x = Alzheimer.s.Diagnosis, y = BMI)) + geom_boxplot() + facet_wrap(~ Physical.Activity.Level)

```
We can see that for both Diet and Physical Activity level, the Alzheimer's Diagnosis occurrence doesn't change much. In other words, BMI exerts about the same effect on diagnosis no matter the dietary habits nor physical activity levels.






### 2 Model Fitting Process 

```{r}

# Convert the specified columns to numeric and combine them into a data frame
new_data <- as.data.frame(lapply(factor_list, function(col) as.numeric(data[[col]])))

# Optionally, set the column names correctly
names(new_data) <- factor_list

# Extract numeric columns from 'data'
numeric_data <- data[, sapply(data, is.numeric)]

# Assuming new_data is already defined (even as an empty data frame with the same number of rows)
new_data <- cbind(new_data, numeric_data)

#Transform the Alzheimer Diagnosis from 1 and 2 to 0 and 1
new_data$Alzheimer.s.Diagnosis <- as.numeric(as.character(new_data$Alzheimer.s.Diagnosis)) - 1

head(new_data)
```

Partition the data so that we can have the train and test set to evaluate the models used 
```{r}
set.seed(123)

# Create a stratified partition for a 75/25 split
train_index <- createDataPartition(data$Alzheimer.s.Diagnosis, p = 0.75, list = FALSE)

# Split the data into training and test sets
train_data_log <- data[train_index, ]
test_data_log  <- data[-train_index, ]

train_data_glm <- new_data[train_index, ]
test_data_glm <- new_data[-train_index, ]

#Transform target column for the logistic data
train_data_log$Alzheimer.s.Diagnosis <- as.numeric(train_data_log$Alzheimer.s.Diagnosis) - 1
test_data_log$Alzheimer.s.Diagnosis <- as.numeric(test_data_log$Alzheimer.s.Diagnosis) - 1

head(train_data_log)
```

###### Conduct the logistic regression 
```{r}
factor_levels <- sapply(train_data_log, function(x) if (is.factor(x)) length(unique(x)) else NA)

log_reg <- glm(Alzheimer.s.Diagnosis ~ . , 
               data = train_data_log, 
               family = binomial)

log_reg.offset <- glm(Alzheimer.s.Diagnosis ~ . - Age, data = train_data_log, offset = log(Age), family = binomial)
```

## SPENCER
###### Conduct Probit regression 
```{r}
# Convert Alzheimer from Yes or No to binary numerical
data$Alzheimer.s.Diagnosis <- as.numeric(data$Alzheimer.s.Diagnosis == "Yes")

probit_reg <- glm(Alzheimer.s.Diagnosis ~ .,
             data = train_data_log,
             family = binomial(link = "probit"))

AIC(log_reg)
AIC(probit_reg)
```
## Logit vs Probit
```{r}
# logit_probs <- predict(log_reg, newdata = test_data_log, type = "response")
# 
# probit_probs <- predict(probit_reg, newdata = test_data_log, type = "response")
# 
# roc_logit <- roc(test_data_log$Alzheimer.s.Diagnosis, logit_probs)
# 
# roc_probit <- roc(test_data_log$Alzheimer.s.Diagnosis, probit_probs)
# 
# plot(roc_logit, col = "blue", main = "ROC: Logit vs Probit")
# lines(roc_probit, col = "red", lty = 2)
# legend("bottomright", legend = c("Logit", "Probit"),
#        col = c("blue", "red"), lty = c(1, 2))
# auc(roc_logit)
# auc(roc_probit)
# 
# plot(logit_probs, probit_probs, 
#      xlab = "Logit Predicted Probabilities", 
#      ylab = "Probit Predicted Probabilities",
#      main = "Logit vs Probit Predictions",
#      col = "darkgreen", pch = 16)
# abline(0, 1, col = "gray", lty = 2)

aic_values <- c(logit = AIC(log_reg), probit = AIC(probit_reg))
barplot(aic_values,
        col = c("steelblue", "salmon"),
        main = "Model Comparison by AIC",
        ylab = "AIC",
        ylim = c(min(aic_values) - 2, max(aic_values) + 2))
text(x = c(0.7, 1.9), y = aic_values + 0.5, 
     labels = round(aic_values, 2))


```

###### Confusion Matrix
```{r}
cutoff <- 0.4141762

# Reduced
pred.train <- predict(probit_reg, type = "response")
class.train <- ifelse(pred.train > cutoff, 1, 0)

confusionMatrix(factor(class.train), factor(train_data_log$Alzheimer.s.Diagnosis), positive = "1")

pred.test <- predict(probit_reg, newdata = test_data_log, type = "response")
class.test <- ifelse(pred.test > cutoff, 1, 0)

confusionMatrix(factor(class.test), factor(test_data_log$Alzheimer.s.Diagnosis), positive ="1")
```


# Binarization
```{r}
binarizer <- dummyVars(~ Country + Physical.Activity.Level + Smoking.Status + Alcohol.Consumption + Depression.Level + Sleep.Quality + Dietary.Habits + Air.Pollution.Exposure + Employment.Status + Marital.Status + Social.Engagement.Level + Income.Level + Stress.Levels,
                       data = data,
                       fullRank = T)
binarized.vars <- data.frame(predict(binarizer, newdata = data))
head(binarized.vars)

data.bin <- cbind(data, binarized.vars)
data.bin$Country <- NULL
data.bin$Physical.Activity.Level <- NULL
data.bin$Smoking.Status <- NULL
data.bin$Alcohol.Consumption <- NULL
data.bin$Depression.Level <- NULL
data.bin$Sleep.Quality <- NULL
data.bin$Dietary.Habits <- NULL
data.bin$Air.Pollution.Exposure <- NULL
data.bin$Employment.Status <- NULL
data.bin$Marital.Status <- NULL
data.bin$Social.Engagement.Level <- NULL
data.bin$Income.Level <- NULL
data.bin$Stress.Levels <- NULL

data.train.bin <- data.bin[train_index, ]
data.test.bin <- data.bin[-train_index, ]

head(data.train.bin)
```


# Fit the new Probit model with binarization
```{r}
probit.full <- glm(Alzheimer.s.Diagnosis ~ .,
                   data = data.train.bin,
                   family = binomial("probit"))
probit.null <- glm(Alzheimer.s.Diagnosis ~ 1,
                   data = data.train.bin,
                   family = binomial("probit"))
probit.reduced <- stepAIC(probit.null,
                          direction = "forward",
                          scope = list(upper = probit.full, lower = probit.null),
                          trace = 0)
# Extract the AIC history from the stepwise selection
aic_values <- probit.reduced$history[, "AIC"]

```

# Step-Wise Graphic
```{r}
stepwise_AIC <- probit.reduced$aic
plot(stepwise_AIC, type = "b", col = "blue", 
     main = "Stepwise Selection: AIC vs. Number of Predictors", 
     xlab = "Step Number", ylab = "AIC", pch = 16, cex = 1.5)
```


###### Summary of reduced model
```{r}
summary(probit.reduced)
```



```{r}
head(data.train.bin)
```

###### Confusion Matrix for Reduced Model
```{r}
cutoff <- 0.4141762
print("REDUCED TRAIN")
pred.train.reduced <- predict(probit.reduced, type = "response")
class.train.reduced <- ifelse(pred.train.reduced > cutoff, 1, 0)

confusionMatrix(factor(class.train.reduced), factor(data.train.bin$Alzheimer.s.Diagnosis), positive = "1")

print("TEST REDUCED")
pred.test.reduced <- predict(probit.reduced, newdata = data.test.bin, type = "response")
pred.test.full <- predict(probit.full, newdata = data.test.bin, type = "response")
pred.test.null <- predict(probit.null, newdata = data.test.bin, type = "response")

class.test.reduced <- ifelse(pred.test.reduced > cutoff, 1, 0)

confusionMatrix(factor(class.test.reduced), factor(data.test.bin$Alzheimer.s.Diagnosis), positive ="1")
```
# ROC's

```{r}
# Create ROC object
roc_reduced <- roc(data.test.bin$Alzheimer.s.Diagnosis, pred.test.reduced)
roc_null <- roc(data.test.bin$Alzheimer.s.Diagnosis, pred.test.null)
roc_full <- roc(data.test.bin$Alzheimer.s.Diagnosis, pred.test.full)

# Plot the ROC curve
plot(roc_reduced, col = "lightblue", main = "ROC Curve - Reduced Model")
lines(roc_null, col = "red", lwd = 2, lty = 2)
lines(roc_full, col = "darkgreen", lwd = 2, lty = 3)
text(0.7, 0.2, paste("AUC (Full) = ", round(auc(roc_full), 4)), col = "green", cex = 0.9)
text(0.7, 0.1, paste("AUC (Null) = ", round(auc(roc_null), 4)), col = "red", cex = 0.9)
text(0.7, 0.05, paste("AUC (Reduced) = ", round(auc(roc_reduced), 4)), col = "blue", cex = 0.9)
legend("bottomright", legend = c("Reduced Model", "Null Model", "Full Model"),
       col = c("blue", "red", "green"), lty = c(1, 2, 3), lwd = 2)
```
## FRANCISCO

##### Plot the Cross Validation for GLM Elastic Net
```{r}
alpha_val <- 0.75

#Assign input values for glmnet compatibility
x_train <- model.matrix(Alzheimer.s.Diagnosis ~ . - 1, data = train_data_glm)
x_test <- model.matrix(Alzheimer.s.Diagnosis ~ . - 1, data = test_data_glm)

y_train <- train_data_glm$Alzheimer.s.Diagnosis
y_test <- test_data_glm$Alzheimer.s.Diagnosis

#Transform to numeric to plot lambda
y_train_numeric <- as.numeric(as.character(y_train))
y_test_numeric <- as.numeric(as.character(y_test))

# Perform cross-validation to search for the optimal lambda
lambda_Search <- cv.glmnet(x_train, y_train_numeric, alpha = alpha_val, nfolds = 10, family = "binomial")

# Plot the cross-validation error versus lambda.
lambda_Search$lambda.min
lambda_Search$lambda.1se
plot(lambda_Search)
```

##### Fit the elastic net logisitc regression model 
```{r}
log_elastic_net <- glmnet(x_train, y_train, family = "binomial", lambda = lambda_Search$lambda.min, alpha = alpha_val)
print(log_elastic_net)
coef_opt <- coef(log_elastic_net, s = lambda_Search$lambda.min)
print(coef_opt)
```

##### Train and Test Accuracy of the Logistic Elastic Net Model 
```{r}
pred_prob_train <- predict(log_elastic_net, newx = x_train, s = lambda_Search$lambda.min, type = "response")
# Convert predicted probabilities to class labels using a threshold of 0.5:
pred_class_train <- ifelse(pred_prob_train >= 0.5, 1, 0)
# Calculate accuracy (percent correct) for the training set:
accuracy_train <- mean(pred_class_train == y_train)
cat("Training Accuracy:", accuracy_train, "\n")

pred_prob_test <- predict(log_elastic_net, newx = x_test, s = lambda_Search$lambda.min, type = "response")
pred_class_test <- ifelse(pred_prob_test >= 0.5, 1, 0)
# Calculate accuracy for the test set:
accuracy_test <- mean(pred_class_test == y_test)
cat("Test Accuracy:", accuracy_test, "\n")
```

##### Train & Test Accuracy of the Logistic Regression Model 
```{r}
# Predict on the training set
train_pred_prob <- predict(log_reg, newdata = train_data_log, type = "response")
train_pred_class <- ifelse(train_pred_prob >= 0.5, 1, 0)
# Extract the actual values (ensure they are numeric 0/1)
train_actual <- as.numeric(as.character(train_data_log$Alzheimer.s.Diagnosis))
# Calculate accuracy for the training set
train_accuracy <- mean(train_pred_class == train_actual)
cat("Training Accuracy:", train_accuracy, "\n")

# Predict on the test set
test_pred_prob <- predict(log_reg, newdata = test_data_log, type = "response")
test_pred_class <- ifelse(test_pred_prob >= 0.5, 1, 0)
# Extract the actual values from the test set
test_actual <- as.numeric(as.character(test_data_log$Alzheimer.s.Diagnosis))
# Calculate accuracy for the test set
test_accuracy <- mean(test_pred_class == test_actual)
cat("Test Accuracy:", test_accuracy, "\n")


# Predict on the Offset data
train_pred_prob <- predict(log_reg.offset, newdata = train_data_log, type = "response")
train_pred_class <- ifelse(train_pred_prob >= 0.5, 1, 0)
# Extract the actual values (ensure they are numeric 0/1)
train_actual <- as.numeric(as.character(train_data_log$Alzheimer.s.Diagnosis))
# Calculate accuracy for the training set
train_accuracy <- mean(train_pred_class == train_actual)
cat("Offset Training Accuracy:", train_accuracy, "\n")

# Predict on the test set
test_pred_prob <- predict(log_reg.offset, newdata = test_data_log, type = "response")
test_pred_class <- ifelse(test_pred_prob >= 0.5, 1, 0)
# Extract the actual values from the test set
test_actual <- as.numeric(as.character(test_data_log$Alzheimer.s.Diagnosis))
# Calculate accuracy for the test set
test_accuracy <- mean(test_pred_class == test_actual)
cat("Offset Test Accuracy:", test_accuracy, "\n")

```


## PCA

```{r}
#Set all non-numeric predictors to NULL
AlzPCA <- data
AlzPCA$Country <- NULL
AlzPCA$Physical.Activity.Level <- NULL
AlzPCA$Smoking.Status <- NULL
AlzPCA$Alcohol.Consumption <- NULL
AlzPCA$Depression.Level <- NULL
AlzPCA$Sleep.Quality <- NULL
AlzPCA$Dietary.Habits <- NULL
AlzPCA$Air.Pollution.Exposure <- NULL
AlzPCA$Employment.Status <- NULL
AlzPCA$Marital.Status <- NULL
AlzPCA$Social.Engagement.Level <- NULL
AlzPCA$Income.Level <- NULL
AlzPCA$Stress.Levels <- NULL
AlzPCA$Gender <- NULL
AlzPCA$Diabetes <- NULL
AlzPCA$Hypertension <- NULL
AlzPCA$Cholesterol.Level <- NULL
AlzPCA$Family.History.of.Alzheimer.s <- NULL
AlzPCA$Genetic.Risk.Factor..APOE.ε4.allele. <- NULL
AlzPCA$Urban.vs.Rural.Living <- NULL
AlzPCA$Alzheimer.s.Diagnosis <- NULL
# Run PCA for all numeric except response
PCA <- prcomp(AlzPCA,
              center = TRUE,
              scale. =  TRUE)

# PC loadings
PCA$rotation
head(PCA$x, 10)
```

### Importance of Each PC
```{r}
summary(PCA)
screeplot(PCA, npcs = 4, type = "lines")  

```

### Create PCA Dataset and Partition
```{r}

# Replace numeric predictors with PCA and add back in the response
AlzPCAdata <- cbind(AlzPCA, PCA$x, data$Alzheimer.s.Diagnosis)
AlzPCAdata$Age <- NULL
AlzPCAdata$Education.Level <- NULL
AlzPCAdata$BMI <- NULL
AlzPCAdata$Cognitive.Test.Score <- NULL
names(AlzPCAdata)[names(AlzPCAdata) == 'data$Alzheimer.s.Diagnosis'] <- 'Alzheimer.s.Diagnosis'

# Partition the new dataset
set.seed(123)
partition <- createDataPartition(AlzPCAdata$Alzheimer.s.Diagnosis, p = 0.75,
                                 list = FALSE)
PCA.Data.Train <- AlzPCAdata[partition, ]
PCA.Data.Test <- AlzPCAdata[-partition, ]

# Ensure a good split was made
mean(PCA.Data.Test$Alzheimer.s.Diagnosis)
mean(PCA.Data.Train$Alzheimer.s.Diagnosis)
```

### Create GLM models with PCA
```{r}
PCAM3 <- glm(Alzheimer.s.Diagnosis ~ . - PC4, data = PCA.Data.Train, family = binomial)
summary(PCAM3)

PCAM4 <- glm(Alzheimer.s.Diagnosis ~., data = PCA.Data.Train, family = binomial)
summary(PCAM4)
```

### PCA Model Performance w/ Confusion Matrices
```{r}
PCA.pred.train <- predict(PCAM4, type = "response")
PCA.class.train <- ifelse(PCA.pred.train > cutoff, 1, 0)

confusionMatrix(factor(PCA.class.train), factor(PCA.Data.Train$Alzheimer.s.Diagnosis), positive = "1")

PCA.pred.test <- predict(PCAM4, newdata = PCA.Data.Test, type = "response")
PCA.class.test <- ifelse(PCA.pred.test > cutoff, 1, 0)

confusionMatrix(factor(PCA.class.test), factor(PCA.Data.Test$Alzheimer.s.Diagnosis), positive = "1")
```

### PCA Model Performance w/ ROC Curve
```{r}
# Create ROC object
roc_PCA <- roc(PCA.Data.Test$Alzheimer.s.Diagnosis, PCA.pred.test)
auc(roc_PCA)

# Plot the ROC curve
plot(roc_PCA, col = "blue", main = "ROC Curve - PCAM4")
text(0.7, 0.2, paste("AUC  = ", round(auc(roc_PCA), 4)), col = "blue", cex = 0.9)

```